Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-05-27 paper/main.tex
#19

%\usepackage[inline]{showlabels}

@@

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \label{fig:related-work:continuity:ktypes}
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{A topological representation of the data (in purple) is a simplified representation of how the records in a dataset are connected. For example, a 1D continuous function is evaluated on a 1D continuous interval, and a map is 2D continuous surface. A network is a union of discrete nodes and continuous lines. The nodes represent the continuity of a table of discrete weather station data and the graph edges are encoded with the distances between the nodes.}
\end{figure}
The topological properties of a dataset are the ways in which elements are organized, as described by Wilkenson\cite{wilkinsonGrammarGraphics2005}. Some examples, as shown in \autoref{fig:related-work:continuity:ktypes} are discrete rows of a table, a continuous 1D line, or a surface. The topological space (in purple) and the data form a key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. For example  in \autoref{fig:related-work:continuity:ktypes}, a point on the green line is mapped to by evaluating the gaussian on a point on the interval and each pixel in the map can be found via a lookup on the 2D surface. This general formulation of topology gives us a language for expressing the continuity of a dataset. The network in \autoref{fig:related-work:continuity:ktypes} is a union of a 0D discrete points and 1D lines connecting those points. The data record in the yellow row can be expressed as a constant function of the yellow network node, while the data on the graph edge between any two points is a function of the distance between those points.

\begin{figure}[h!]
  \label{fig:related-work:continuity:borked}
  \includegraphics[width=1\columnwidth]{continuity.png}
  \caption{The \texttt{line} and \texttt{image} visual algorithms take in the same array of data. The \texttt{line} assumes that the array is a list of lines, while the \texttt{image} assumes that the array is an image.}
\end{figure}
The continuity matters to visualization libraries because visual algorithms, such as a the line plot and image plot in \autoref{fig:related-work:continuity:borked}, assume the continuity of their data, as described in taxonomies of visualization algorithms by Troy, M\"{o}ller, and Chi \cite{toryRethinkingVisualizationHighlevel2004,chiTaxonomyVisualizationTechniques2000}. In \autoref{fig:related-work:continuity:borked}, both \texttt{line} and \textt{image} algorithms are fed the same input matrix. The line plot assumes this matrix is a list of lines and plots it as such. The image map assumes that the data is an image and plots it as such. Neither function has a way of asking the matrix what it is, nor does the matrix encode information about the continuity of the data it contains. While this work is not about choosing the right visualization type for the given continuity, the ability to encode the continuity expectations in the visual algorithm and the data container would provide a way of ensuring that the assumptions match.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fbtau.png}
  \caption{A non-trivial bundle provides a way of expressing data where the fiber changes along the indexing base space. Here $\dfiber_{\dbasepoint_1}$ and $\dfiber_{\dbasepoint_2}$ go in reverse directions \note{do fibers have to be isomorphic?}. The global section $sin$ returns $sin(\dbasepoint_{i})$ which is a value between $[0,1]$
  \label{fig:atct:stripbundle}}
\end{figure}

@@

We also use fiber bundles to represent the output of a visualization algorithm, which we term the graphic but generalizes to output on any display space, such as a screen or 3D print. 
\begin{equation}
  \label{eq:atct:fb:graphic}
  \gfiberc \hookrightarrow \gtotalc \xrightarrow{\pi} \gbasec
\end{equation}
The base space \gbasec\ is a parameterization of the display area, for example the inked bounding box in cairo \cite{CairographicsOrg}. The fiber space \gfiberc\ is an abstraction of the renderer fields, for example $\{x,\,y,\,r,\,g,\,b,\,a\}$. The sections of the graphic 
\begin{equation}
  \label{eq:atct:fb_graphic_section}
  \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}} \coloneqq \big\{\gsectionc: \opensetgc\rightarrow \gtotalc\restriction_{\opensetgc} \; \bigm{\vert} \pi(\gsectionc(\gbasepointc)) = \gbasepointc\;for\, all\; \gbasepointc \in \opensetgc \big\}
\end{equation}
are functions that generate different graphics, for example markers for a scatter plot, bars for a bar chart, cells of a heatmap, or pixels of an image.

\begin{figure}[h!]
  \label{fig:atct:fb:graphic}
  \includegraphics[width=1\columnwidth]{fb_rho.png}
  \caption{A bundle representation of a section that generates a scatter marker on a 2D screen. The fiber space encodes the potential color and position of each pixel in a prerender space. The marker is a 2d graphic; therefore the base space that parameterizes the graphic area \dbasec\ is a 2d disc. This section \gsection\ describes how to build a square red marker; a section evaluated on a single point $\gbasepoint \in \gbase$ is an infinitesimally small piece of the square, here illustrated as the small piece of the square outlined in green.}
\end{figure}

In \autoref{fig:atct:fb:graphic}, the section function $\gsection$ encodes the drawing of a red square marker. The renderer subsamples the position fibers  \textit{x} and \textit{y} of the graphic fiber space \gfiber. The preimage of the section at these coordinates returns a set of coordinates \({\gsection^{-1}_{x,y}} = \{\gbasepoint | \gsection(\gbasepoint)\restriction_{x,y} = \(x,y\)\}\) at the points \(\gbasepoint \in \gbase\) in the graphic base topological space. The section is then evaluated at these points \(\gsection\restriction_{\gsection^{-1}_{x,y}}\) to return a set of points in the fiber. One of these points, enlarged for visibility, is illustrated as the small square swatch inside the larger marker in \autoref{fig:atct:fb}. The renderer is handed some version of $\gsection$ to produce the final graphic.

\subsection{Encoding Topology and Field Types}
\label{sec:atct:io}
As described in \autoref{sec:atct:fiber-bundles}, we model data as a map from base space to fiber space.To encapsulate the structure carried in these maps, and provide a way to define these maps in reusable and composable manner, we define categories of base and fiber spaces.

\begin{definition}
  \label{sec:atct:io:base}
  The base space category $\mathcal{\opensetc}$ has open set objects $\opensetc$ and inclusion morphisms $\iota: \opensetc_1 \rightarrow \opensetc_2$ such that $\opensetc_1 \subseteq \opensetc_2$. 
\end{definition}

@@

A sheaf is a presheaf that satisfies the locality and gluing axioms \cite{bakerMathsSheaf}. 
\begin{definition}
\label{def:atct:sheaf:locality}
The locality axiom is that given a union of opensets $\mathscr{\opensetc} = \bigcup\limits_{i\in I} \opensetc_i$ and $\dsectionc^{a}, \dsectionc^{b} \in \sheafc(\mathscr{\opensetc})$,  if $\dsectionc^{a}\restriction_{\opensetc_i} = \dsectionc^{b}\restriction_{\opensetc_i}$ for each $\opensetc_i \in \opensetc$ then $\dsectionc^{a} = \dsectionc^{b}$
\end{definition}
This means that two sections in a sheaf are equal when they evaluate to the same values over all open sets in a collection. 
\begin{definition}
\label{def:atct:sheaf:gluing}
The gluing axiom is that given $\dsectionc^{i} \in \sheafc(\opensetc_i)$ such that ${\dsectionc^{i}}\restriction_{\opensetc_i\cap \opensetc_j} = {\dsectionc^{j}}\restriction_{\opensetc_i \cap \opensetc_j}$ for $\opensetc_i, \opensetc_j \in \mathscr{\opensetc}$, there exists $\dsectionc \in \sheafc(\mathscr{\opensetc})$ such that $\dsectionc\restriction_{\opensetc_i} = {\dsectionc^{i}}$. 
\end{definition}
This means we can construct a section $\dsection$ on the collection of open sets that is the union of sections defined on specific open sets. This expresses an equivalency between a concatenated representation of data and the distributed representation of the same data. In this model, we represent data and graphics as a sheaf; doing so allows us to express that data sources are expected to implement the bookkeeping that a sheaf requires, meaning inclusions, restrictions, locality, and gluing must be preserved.

@@

\begin{figure}[h!]
  \label{fig:atct:phi}
  \includegraphics[width=\columnwidth]{phi.png}
  \caption{Values in a data set can be transformed in three ways: $\dfunch$-values can change position, .e.g transposed;  $\dfunct$-values can change, e.g. doubled; $\dfunc$ - values can change position and value}
\end{figure}
\autoref{fig:atct:phi} provides an example of a base space change \dfunch, a fiber space change \dfunct, and a composition of the two \dfunc\ applied to each data point $x_{\dbasepoint} \in \texttt{data}$. As defined in \autoref{eq:atct:morphism:verify_base}, 
since transposition is a $(\dfunch, \dfuncpull)$, every data value $x_{\dbasepoint}$ must be equal to the value at the corresponding location in the transposed matrix $x_{\dbasepoint} = x_{\dbasepoint^{\prime}}, \dfunch(\dbasepoint^{\prime}) = \dbasepoint$ for all $x_{\dbasepoint^{\prime}} \in texttt{data.T}$ . In \autoref{fig:atct:phi}, this is demonstrated in the columns becoming rows such that there are no new values and each value retains the same neighboring values. As defined in \autoref{eq:atct:morphism:fiber} scaling is a transform $\dfunct$ that changes the value of each point but not its location. In \autoref{fig:atct:phi}, this means that $x_{\dbasepoint^{\prime}}= 2*x_{\dbasepoint}, \dbasepoint = \dbasepoint_{\prime}$ for all indices in \texttt{data}, \texttt{data.T}. As defined in \autoref{eq:atct:morphism:all} and illustrated in \autoref{fig:atct:phi}, these transformations can combine to yield a matrix that is both transposed and scaled such that $x_{\dbasepoint^{\prime}} = 2*x_{\dbasepoint}$ for all  
$\dfunch(\dbasepoint^{\prime}) = \dbasepoint$

@@

\begin{figure}[h!]
  \label{fig:atct:morphisms:sheaf}
  \includegraphics*[width=1\columnwidth]{xi_scatter.png}
  \caption{The base space functor $\vindexc$ is the mapping from each disc $\gbasec_i$ to it's corresponding point $\dbasepointc_i$. Over each point is a data record $\dsectionc$. The pullback function $\vindexpullc$ copies these records over $\gbase$ so that each piece of each marker knows the data associated with it. Each graphic disc has a corresponding graphic function $\gsectionc$. The pushforward function $\vindexpushc$ converts $\gsectionc$ into the visual parameters associated with each data point $\dbasepoint$.}
\end{figure}

@@

We propose that visualization libraries implement morphisms from a data sheaf $\sheafc_{\dtotal, \dbase}$ to a graphic sheaf $\sheafc_{\gtotal, \gbase}$ 
\begin{equation}
  \label{eq:atct:sheaves:homset}
  \begin{tikzcd}
    \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} 
    \arrow[dd, "\textcolor{set}{Hom}_{\sheafc_{\dbasec}}"', color=homset] 
    \arrow[rrdd, "\textcolor{set}{Hom}_{\sheafc_{\dbasec},\sheafc_{\gbasec}}", color=homset] 
    \arrow[rr, "\vindexpullc", color=functor] &  &
    \cgamma{\opensetgc}{\vindexpullc\dtotalc\restriction_{\opensetgc}} 
    \arrow[dd, "\textcolor{set}{Hom}_{\sheafc_{\gbasec}}", color=homset] \\
     & & \\
    \cgamma{\opensetc}{\vindexpushc\gtotalc\restriction_{\opensetc}} &  & 
    \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}} 
    \arrow[ll, "\vindexpushc"', color=functor]                  
    \end{tikzcd}
\end{equation}
and the hom set is the set of all morphisms from one sheaf to the other. The functors $\vindexpullc, \vindexpushc$ are adjoint, meaning that there is a functorial isomorphism \cite{harder2008lectures} such that 
\begin{equation}
  \label{eq:atct:sheaves:homset:functors}
\begin{split}
  & \textcolor{set}{Hom}_{\sheafc_{\gbasec}}(\vindexpullc\sheafc_{\dtotalc, \dbasec},\sheafc_{\gtotalc,\gbasec})\\
  \simeq & \textcolor{set}{Hom}_{\sheafc_{\dbasec}}(\sheafc_{\dtotalc, \dbasec},\vindexpushc\sheafc_{\gtotalc,\gbasec}) \\
  \simeq & \textcolor{set}{Hom}_{\sheafc_{\dbasec},\sheafc_{\gbasec}}(\sheafc_{\dtotalc, \dbasec},\sheafc_{\gtotalc,\gbasec}) \\
\end{split} 
\end{equation}
which means that the diagram of sheaf morphisms \autoref{eq:atct:sheaves:homset} is commutative; therefore the functors $\vindexpullc$ and $\vindexpushc$ can be used to adapt morphisms written over one space to another space. 
This provides us flexibility over which space to construct data to graphic transforms. For example, specifications such as svg\cite{quintScalable2003} and Vega\cite{satyanarayanDeclarativeInteractionDesign2014}, typically define their transformations in data space and return specifications for the render to evaluate. An artist may also want to specify the transformations in a proxy of graphic space, for example to generate on-demand dynamically resampled continuous lines.

\section{Artist: Data to Graphic}
\label{sec:artist}
In this work we propose that visualization libraries are implementing a subset of the functions in the hom sets in \autoref{eq:atct:sheaves:homset}. We call these subset of functions the artist:
\begin{align}
  \label{eq:artist:hom_transport}
  \vartistc:& \cgamma{\dbasec}{\dtotalc}\textcolor{artist}{\rightarrow} \cgamma{\gbasec}{\gtotalc} \in \textcolor{set}{Hom}_{\sheafc_{\dbase}, \sheafc_{\gbase}}\\
  \vartistc^{\dbasec}:&\cgamma{\dbasec}{\dtotalc}  \textcolor{artist}{\rightarrow} \cgamma{\dbase}{\vindexpushc\gtotalc} \in \textcolor{set}{Hom}{\sheafc_{\dbase}}\\
  \vartistc^{\gbasec}:& \cgamma{\gbasec}{\vindexpullc\dtotalc} \textcolor{artist}{\rightarrow} \cgamma{\gbasec}{\gtotalc} \in \textcolor{set}{Hom}_{\sheafc_{\gbase}}
\end{align}
 Because the artists can be constructed as morphisms of sheave over the same base spaces, as listed in \autoref{eq:atct:sheaves:homset:functors}, they are natural transformations
\begin{align}
  \label{eq:artist:natural_transform}
  \vartistc^{\dbasec}:& \sheafc_{\dbasec, \dtotalc} \textcolor{artist}{\Rightarrow} \vindexpushc \sheafc_{\gbasec, \gtotalc}\\
  \vartistc^{\gbasec}:&\vindexpullc\sheafc_{\dbasec, \dtotalc} \textcolor{artist}{\Rightarrow}\sheafc_{\gbasec, \gtotalc}
\end{align}
which means that they are maps of functors that take the same input object and return objects in the same category\cite{milewskiCategoryTheoryProgrammers}. As illustrated in \autoref{eq:atct:sheaves:homset}, the sheaf functors
\begin{equation}
  \label{eq:artist:sheaf:base}
    \begin{tikzcd}
      \cgamma{\dbasec}{\dtotalc} &  & \dbasec \arrow[ll, "{\sheafc_{\dbasec, \dtotalc}}"', maps to, color=sheaf] \arrow[rr, "{\vindexpushc\sheafc_{\gbasec, \gtotalc}}", maps to, color=sheaf] &  & \cgamma{\dbase}{\vindexpushc\gtotalc} 
      \end{tikzcd}
\end{equation}
take as input the same object \opensetc\ and return sets of data and graphic sections that are objects in \setb. As a map between these sheaf functors, the artist has to preserve the $\iota, \iota^*$ morphisms of the presheaf functor, described in \autoref{eq:atct:presheaf}, such that the following diagram commutes

@@

An artist is equivariant when changes to expected and actual are equivariant. As introduced in \autoref{eq:atct:morphism:base}, the base space transformation \dfunch\ is invariant because $\dsection\restriction_{\openset} = \dsection\restriction_{\dfunch(\openset^{\prime})}$. This means that, for all points in the data $\dbasepoint \in \dbase$, the measurement should not change if only the base space is transformed 
\begin{equation}
  \label{eq:atct:equivarance:verify:base}
  \equivc(\dsectionc)(\dfunch(\dbasepointc^{\prime})) = \extractmc(\vartistc(\dsectionc))(\dbasepointc)
\end{equation}
On the other hand, a change in sections \autoref{eq:atct:morphism:fiber} induces an equivalent change in measurements
\begin{equation}
  \label{eq:atct:equivarance:verify:fiber}
  \equivc(\dfunctc(\dsectionc))(\dbasepoint) = \dfunctc_{\measure}(\extractmc(\vartistc(\dsectionc))(\dbasepoint))
\end{equation}
The change in measurements $\dfunctc_{\measure}$ is defined by the developer as the symmetry between data and graphic that the artist is expected to preserve.

\begin{figure}[h!]
  \label{fig:artist:equivariance}
  \includegraphics[width=1\columnwidth]{equivariance.png}
  \caption{This artist is equivariant because the input data and output graphic representation are transposed $\dfunch$, scaled $\dfunct$ and both transposed and scaled $\dfunc$ in the same manner.}
\end{figure}
For example, in \autoref{fig:artist:equivarance}, the measurable variable is color. This is a visual representation of the data shown in \autoref{fig:atct:phi}, and as such the equivariant transformations are an equivalent transposition and scaling of the colors. This visualization is equivariant with respect to base space transformations, as defined in \autoref{eq:atct:equivarance:verify:base}, because the color values at the new position at the old position $measure_\dbasepoint^{\prime} = \measure_{\dbasepoint}$. This visualization is also equivariant with respect to fiber wise transformations, as defined in \autoref{eq:atct:equivarance:verify:fiber}, because the colors are consistently scaled in the same was the data. For example, the values that have become 2 and 4 in the $\dfunctc$ and $\dfunc$ panels are colored the same as the original 2 and 4 values in the first panel. The equivariance in this visualization is composable, as shown in the colors being both transposed and scaled correctly in the $\dfunc$ panel.

>>>>

\usepackage[inline]{showlabels}

@@

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{A topological representation of the data (in purple) is a simplified representation of how the records in a dataset are connected. For example, a 1D continuous function is evaluated on a 1D continuous interval, and a map is 2D continuous surface. A network is a union of discrete nodes and continuous lines. The nodes represent the continuity of a table of discrete weather station data and the graph edges are encoded with the distances between the nodes. \label{fig:related-work:continuity:ktypes}}
\end{figure}
The topological properties of a dataset are the ways in which elements are organized, as described by Wilkenson\cite{wilkinsonGrammarGraphics2005}. Some examples, as shown in \autoref{fig:related-work:continuity:ktypes} are discrete rows of a table, a continuous 1D line, or a surface. The topological space (in purple) and the data form a key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. For example  in \autoref{fig:related-work:continuity:ktypes}, a point on the green line is mapped to by evaluating the gaussian on a point on the interval and each pixel in the map can be found via a lookup on the 2D surface. This general formulation of topology gives us a language for expressing the continuity of a dataset. The network in \autoref{fig:related-work:continuity:ktypes} is a union of a 0D discrete points and 1D lines connecting those points. The data record in the yellow row can be expressed as a constant function of the yellow network node, while the data on the graph edge between any two points is a function of the distance between those points.

\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{continuity.png}
  \caption{The \texttt{line} and \texttt{image} visual algorithms take in the same array of data. The \texttt{line} assumes that the array is a list of lines, while the \texttt{image} assumes that the array is an image.  \label{fig:related-work:continuity:borked}}
\end{figure}
The continuity matters to visualization libraries because visual algorithms, such as a the line plot and image plot in \autoref{fig:related-work:continuity:borked}, assume the continuity of their data, as described in taxonomies of visualization algorithms by Troy, M\"{o}ller, and Chi \cite{toryRethinkingVisualizationHighlevel2004,chiTaxonomyVisualizationTechniques2000}. In \autoref{fig:related-work:continuity:borked}, both \texttt{line} and \textt{image} algorithms are fed the same input matrix. The line plot assumes this matrix is a list of lines and plots it as such. The image map assumes that the data is an image and plots it as such. Neither function has a way of asking the matrix what it is, nor does the matrix encode information about the continuity of the data it contains. While this work is not about choosing the right visualization type for the given continuity, the ability to encode the continuity expectations in the visual algorithm and the data container would provide a way of ensuring that the assumptions match.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fbtau.png}
  \caption{A non-trivial fiber bundle is bundle where the (by definition isomorphic) fibers change along the indexing base space. Here $\dfiber_{\dbasepoint_1}$ and $\dfiber_{\dbasepoint_2}$ go in reverse directions and the global section $sin$ returns $sin(\dbasepoint_{i})$ which is a value between $[0,1]$
  \label{fig:atct:stripbundle}}
\end{figure}

@@

\subsection{Uniform Abstract Graphic Representation}
Fiber bundles are such a generalizable data abstraction that we can also use them to encode the output of a visual bundle; bundles allow for fully expressing the properties of the graphic that must be symmetric to the data in an equivariant (\autoref{sec:related-work:equivariance}) data to graphic transformation. We denote the output as a graphic, but the use of bundles allows us to generalize to output on any display space, such as a screen or 3D print. 
\begin{equation}
  \label{eq:atct:fb:graphic}
  \gfiberc \hookrightarrow \gtotalc \xrightarrow{\pi} \gbasec
\end{equation}
\begin{equation}
  \label{eq:atct:fb_graphic_section}
  \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}} \coloneqq \big\{\gsectionc: \opensetgc\rightarrow \gtotalc\restriction_{\opensetgc} \; \bigm{\vert} \pi(\gsectionc(\gbasepointc)) = \gbasepointc\;for\, all\; \gbasepointc \in \opensetgc \big\}
\end{equation}
The base space \gbasec\ is a parameterization of the display area, for example the inked bounding box in cairo \cite{CairographicsOrg}. The fiber space \gfiberc\ is an abstraction of the renderer fields, for example $\{x,\,y,\,r,\,g,\,b,\,a\}$. The sections $\gsection$ of the graphic are functions that go from a point in the graphic base to elements in a prerendering space.

\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{fb_rho.png}
  \caption{For a 2D display a section $\gsection$ maps into a fiber $\dfiber$ that encodes an RGB prerender space with infinite resolution $\mathcal{R}^{2}$. Each tiny colored box is an approximation of the return value of the same section function $\gsection$ evaluated on different points $\gbasepoint \in \gbase$ in the base space. \label{fig:atct:fb:graphic}}
\end{figure}

In \autoref{fig:atct:fb:graphic}, the section function $\gsection$ maps into the fiber for a simplified 2D RGB infinite resolution prerender space and returns the $\{x,y,r,g,b\}$ values of a pixel in an infinite resolution space. In \autoref{fig:atct:fb:graphic} these pixels are approximated as the small orange and black colored boxes. Each box is the output of the $\gsection(\gbasepoint)$ section that intersects the box. The set of all pixels returned by a section evaluated on a given visual base space $\gsection|_{\gbase}$ can yield a visual element, such as a marker, line, or piece of a glyph.

\subsection{Topology and Field Categories}
\label{sec:atct:io}
We define categorical types for the topology and the fibers because categories are defined both by the objects of the category and the valid morphisms on the object. This allows for the formal construction of an equivariance aware type that holds the topology, field, type, and what changes to either must be preserved.

\begin{definition}
  \label{sec:atct:io:base}
  The base space category $\mathcal{\dbasec}$ has open set objects $\opensetc$ and inclusion morphisms $\iota: \opensetc_1 \rightarrow \opensetc_2$ such that $\opensetc_1 \subseteq \opensetc_2$. 
\end{definition}

@@

The conditions on sheaves express the presheaf properties of continuity preservation as applied to sections, which as mentioned in \autoref{sec:atct:fiber-bundles} are how data is modeled in this framework. A sheaf is a presheaf that satisfies the locality and gluing axioms \cite{bakerMathsSheaf}. 
\begin{definition}
\label{def:atct:sheaf:locality}
The locality axiom is that given a union of opensets $\mathscr{\opensetc} = \bigcup\limits_{i\in I} \opensetc_i$ and $\dsectionc^{a}, \dsectionc^{b} \in \sheafc(\mathscr{\opensetc})$,  if $\dsectionc^{a}\restriction_{\opensetc_i} = \dsectionc^{b}\restriction_{\opensetc_i}$ for each $\opensetc_i \in \opensetc$ then $\dsectionc^{a} = \dsectionc^{b}$
\end{definition}
This means that two sections in a sheaf are equal when they evaluate to the same values over all open sets in a collection. 
\begin{definition}
\label{def:atct:sheaf:gluing}
The gluing axiom is that given $\dsectionc^{i} \in \sheafc(\opensetc_i)$ such that ${\dsectionc^{i}}\restriction_{\opensetc_i\cap \opensetc_j} = {\dsectionc^{j}}\restriction_{\opensetc_i \cap \opensetc_j}$ for $\opensetc_i, \opensetc_j \in \mathscr{\opensetc}$, there exists $\dsectionc \in \sheafc(\mathscr{\opensetc})$ such that $\dsectionc\restriction_{\opensetc_i} = {\dsectionc^{i}}$. 
\end{definition}
This means we can construct a section $\dsection$ on the collection of open sets that is the union of sections defined on specific open sets. This expresses an equivalency between a concatenated representation of data and the distributed representation of the same data. In this model, we represent data and graphics as a sheaf; doing so allows us to express that data sources are expected to implement the bookkeeping that a sheaf requires, meaning inclusions, restrictions, locality, and gluing must be preserved.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{phi.png}
  \caption{Values in a data set can be transformed in three ways: $\dfunch$-values can change position, .e.g transposed;  $\dfunct$-values can change, e.g. doubled; $\dfunc$ - values can change position and value  \label{fig:atct:phi}}
\end{figure}
\autoref{fig:atct:phi} provides an example of a base space change \dfunch, a fiber space change \dfunct, and a composition of the two \dfunc\ applied to each data point $x_{\dbasepoint} \in \texttt{data}$. As defined in \autoref{eq:atct:morphism:verify_base}, 
since transposition is a $(\dfunch, \dfuncpull)$, every data value $x_{\dbasepoint}$ must be equal to the value at the corresponding location in the transposed matrix $x_{\dbasepoint} = x_{\dbasepoint^{\prime}}, \dfunch(\dbasepoint^{\prime}) = \dbasepoint$ for all $x_{\dbasepoint^{\prime}} \in texttt{data.T}$ . In \autoref{fig:atct:phi}, this is demonstrated in the columns becoming rows such that there are no new values and each value retains the same neighboring values. As defined in \autoref{eq:atct:morphism:fiber} scaling is a transform $\dfunct$ that changes the value of each point but not its location. In \autoref{fig:atct:phi}, this means that $x_{\dbasepoint^{\prime}}= 2*x_{\dbasepoint}, \dbasepoint = \dbasepoint_{\prime}$ for all indices in \texttt{data}, \texttt{data.T}. As defined in \autoref{eq:atct:morphism:all} and illustrated in \autoref{fig:atct:phi}, these transformations can combine to yield a matrix that is both transposed and scaled such that $x_{\dbasepoint^{\prime}} = 2*x_{\dbasepoint}$ for all  
$\dfunch(\dbasepoint^{\prime}) = \dbasepoint$

@@

\begin{figure}[h!]
  \includegraphics*[width=1\columnwidth]{xi_scatter.png}
  \caption{The base space functor $\vindexc$ is the mapping from each disc $\gbasec_i$ to it's corresponding point $\dbasepointc_i$. Over each point is a data record $\dsectionc$. The pullback function $\vindexpullc$ copies these records over $\gbase$ so that each piece of each marker knows the data associated with it. Each graphic disc has a corresponding graphic function $\gsectionc$. The pushforward function $\vindexpushc$ converts $\gsectionc$ into the visual parameters associated with each data point $\dbasepoint$.\label{fig:atct:morphisms:sheaf}}
\end{figure}

@@

We propose that visualization libraries implement morphisms from a data sheaf $\sheafc_{\dtotal, \dbase}$ to a graphic sheaf $\sheafc_{\gtotal, \gbase}$ 
\begin{equation}
  \label{eq:atct:sheaves:homset}
  \begin{tikzcd}
    \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}} 
    \arrow[dd, "\textcolor{set}{Hom}_{\sheafc_{\dbasec}}"', color=homset] 
    \arrow[rrdd, "\textcolor{set}{Hom}_{\sheafc_{\dbasec},\sheafc_{\gbasec}}", color=homset] 
    \arrow[rr, "\vindexpullc", color=functor] &  &
    \cgamma{\opensetgc}{\vindexpullc\dtotalc\restriction_{\opensetgc}} 
    \arrow[dd, "\textcolor{set}{Hom}_{\sheafc_{\gbasec}}", color=homset] \\
     & & \\
    \cgamma{\opensetc}{\vindexpushc\gtotalc\restriction_{\opensetc}} &  & 
    \cgamma{\opensetgc}{\gtotalc\restriction_{\opensetgc}} 
    \arrow[ll, "\vindexpushc"', color=functor]                  
    \end{tikzcd}
\end{equation}
and the hom set is the set of all morphisms from one sheaf to the other. The functors $\vindexpullc, \vindexpushc$ are adjoint, meaning that there is a functorial isomorphism \cite{harder2008lectures} such that 
\begin{equation}
  \label{eq:atct:sheaves:homset:functors}
   \textcolor{set}{Hom}_{\sheafc_{\gbasec}}(\vindexpullc\sheafc_{\dtotalc, \dbasec},\sheafc_{\gtotalc,\gbasec})
  \simeq  \textcolor{set}{Hom}_{\sheafc_{\dbasec}}(\sheafc_{\dtotalc, \dbasec},\vindexpushc\sheafc_{\gtotalc,\gbasec}) 
\end{equation}
such that both are isomorphic to  $\simeq  \textcolor{set}{Hom}_{\sheafc_{\dbasec},\sheafc_{\gbasec}}(\sheafc_{\dtotalc, \dbasec},\sheafc_{\gtotalc,\gbasec})$. This set of ismorphic homsets means that the diagram of sheaf morphisms \autoref{eq:atct:sheaves:homset} is commutative; therefore the functors $\vindexpullc$ and $\vindexpushc$ can be used to adapt morphisms written over one space to another space. 
\note{maybe move to construction}
This provides us flexibility over which space to construct data to graphic transforms. For example, specifications such as svg\cite{quintScalable2003} and Vega\cite{satyanarayanDeclarativeInteractionDesign2014}, typically define their transformations in data space and return specifications for the render to evaluate. An artist may also want to specify the transformations in a proxy of graphic space, for example to generate on-demand dynamically resampled continuous lines.

\section{Artist: Data to Graphic}
\label{sec:artist}
In this work we propose that visualization libraries are implementing a subset of the functions in the hom sets in \autoref{eq:atct:sheaves:homset}. We call these subset of functions the artist:
\begin{align}
  \label{eq:artist:hom_transport}
  \vartistc:& \cgamma{\dbasec}{\dtotalc}\textcolor{artist}{\rightarrow} \cgamma{\gbasec}{\gtotalc} \in \textcolor{set}{Hom}_{\sheafc_{\dbase}, \sheafc_{\gbase}}
\end{align}
 Because the artists can be constructed as morphisms of sheave over the same base spaces, as listed in \autoref{eq:atct:sheaves:homset:functors}, they are natural transformations, which means that they are maps of functors that take the same input object and return objects in the same category\cite{milewskiCategoryTheoryProgrammers}. As illustrated in \autoref{eq:atct:sheaves:homset}, the sheaf functors
\begin{equation}
  \label{eq:artist:sheaf:base}
    \begin{tikzcd}
      \cgamma{\dbasec}{\dtotalc} &  & \dbasec \arrow[ll, "{\sheafc_{\dbasec, \dtotalc}}"', maps to, color=sheaf] \arrow[rr, "{\vindexpushc\sheafc_{\gbasec, \gtotalc}}", maps to, color=sheaf] &  & \cgamma{\dbase}{\vindexpushc\gtotalc} 
      \end{tikzcd}
\end{equation}
take as input the same object \opensetc\ and return sets of data and graphic sections that are objects in \setb. As a map between these sheaf functors, the artist has to preserve the $\iota, \iota^*$ morphisms of the presheaf functor, described in \autoref{eq:atct:presheaf}, such that the following diagram commutes

@@

An artist is equivariant when changes to expected and actual are equivariant. As introduced in \autoref{eq:atct:morphism:base}, the base space transformation \dfunch\ is invariant because $\dsection\restriction_{\openset} = \dsection\restriction_{\dfunch(\openset^{\prime})}$. This means that, for all points in the data $\dbasepoint \in \dbase$, the measurement should not change if only the base space is transformed 
\begin{equation}
  \label{eq:atct:equivarance:verify:base}
  \equivc(\dsectionc)(\dfunchc(\dbasepointc^{\prime})) = \extractmc(\vartistc(\dsectionc))(\dbasepointc)
\end{equation}
On the other hand, a change in sections \autoref{eq:atct:morphism:fiber} induces an equivalent change in measurements
\begin{equation}
  \label{eq:atct:equivarance:verify:fiber}
  \equivc(\dfunctc(\dsectionc))(\dbasepointc) = \dfunctc_{\measure}(\extractmc(\vartistc(\dsectionc))(\dbasepoint))
\end{equation}
The change in measurements $\dfunctc_{\measure}$ is defined by the developer as the symmetry between data and graphic that the artist is expected to preserve.

\begin{figure}[h!]
  \includegraphics[width=1\columnwidth]{equivariance.png}
  \caption{This artist is equivariant because the input data and output graphic representation are transposed $\dfunch$, scaled $\dfunct$ and both transposed and scaled $\dfunc$ in the same manner. \label{fig:artist:equivariance}}
\end{figure}
For example, in \autoref{fig:artist:equivarance}, the measurable variable is color. This is a visual representation of the data shown in \autoref{fig:atct:phi}, and as such the equivariant transformations are an equivalent transposition and scaling of the colors. This visualization is equivariant with respect to base space transformations, as defined in \autoref{eq:atct:equivarance:verify:base}, because the color values at the new position at the old position $measure_\dbasepoint^{\prime} = \measure_{\dbasepoint}$. This visualization is also equivariant with respect to fiber wise transformations, as defined in \autoref{eq:atct:equivarance:verify:fiber}, because the colors are consistently scaled in the same was the data. For example, the values that have become 2 and 4 in the $\dfunctc$ and $\dfunc$ panels are colored the same as the original 2 and 4 values in the first panel. The equivariance in this visualization is composable, as shown in the colors being both transposed and scaled correctly in the $\dfunc$ panel.
